Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-09-10 inst/manuscript/manuscript.tex
some updates

\Abstract{
Generalized linear models (GLMs) describe a non-normal response variable
that may be binary, count, skewed, or a proportion. Typically,
observations in a GLM are assumed independent of one another. For
spatial data, this independence assumption is impractical, as nearby
locations tend to be more similar than locations far apart. The
\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
incorporate spatial correlation (i.e., spatial generalized linear
models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
application of the Laplace approximation via \code{spglm()} for
point-referenced data or \code{spgautor()} for areal (i.e., lattice),
data. \code{spglm()} and \code{spgautor()} closely resemble {glm} from
base \proglang{R} but include arguments that control the spatial
correlation structure. \pkg{spmodel} has many helper functions for model
inspection and diagnostics, some of which leverage other \proglang{R}
packages like {broom} and {emmeans}. \pkg{spmodel} has tools to make
predictions of the latent spatial-mean process at unobserved locations.
\pkg{spmodel} also provides many advanced features like accommodating
geometric anisotropy and nonspatial random effects, simulating spatially
autocorrelated data, and more. Here we use \pkg{spmodel} to illustrate
the modeling of binary, count, skewed and proportion response variables
from several point-referenced and areal data sets.
}

@@

\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm, bbm}
\usepackage{lineno}
\usepackage{caption, subcaption}

@@

@@

In practice, non-Gaussian data are ubiquitous. Non-Gaussian that belong
to an exponential family data can be naturally modeled using a
generalized linear model (GLM) regression framework
\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
In a GLM, an \(n \times 1\) response variable \(\mathbf{y}\) belongs to
a statistical distribution (e.g., Poisson, Binomial) with some mean and
variance. Often, the analysis goal is to study the impact of a linear
function of several explanatory variables on \(\text{y}\) through a GLM.
In this context, the latent (i.e., unobserved) mean of \(\mathbf{y}\),
\(\boldsymbol{\mu}\), is linked to these explanatory variables via a
link function: \begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation} where for a sample size \(n\), \(f(\cdot)\) is a link
function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
\(\mathbf{X}\) is the \(n \times p\) design matrix of explanatory
variables, and \(\boldsymbol{\beta}\) is the \(p \times 1\) vector of
fixed effects. While the mean is typically constrained in some way
(e.g., between zero and one if a probability), the link function
generally makes \(\mathbf{w}\) is unconstrained. Common link functions
inlude the log odds (i.e., logit) link for binary and proportion data
and the log link count and skewed data. Equation\(~\)\ref{eq-glm} can
also be written in terms of the inverse link function,
\(f^{-1}(\cdot)\): \begin{equation}\label{eq-glm2}
\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
\end{equation}

@@

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
In spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, leading to positive
spatial covariance among observations. The consequences of ignoring
spatial covariance in statistical models for spatial data can be severe
and include imprecise parameter estimates as well as misleading standard
errors that inflate Type-I error rates and decrease power
\citep{zimmerman2024spatial}.

An approach for handling spatial data using a GLM is to assume
\(\mathbf{w}\) has spatial covariance. This is achieved by adding to
Equation\(~\)\ref{eq-glm} two random effects, \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\). The random effect \(\boldsymbol{\tau}\) is an
\(n \times 1\) column vector of spatially dependent random errors. We
assume that \(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\text{E}(\cdot)\) and \(\text{Cov}(\cdot)\) denote expectation and
covariance, respectively. The variance parameter \(\sigma^2_\tau\)
controls the magnitude of spatial covariance and is often called a
partial sill, while the matrix \(\mathbf{R}\) is an \(n \times n\)
spatial correlation matrix that depends on a range parameter which
controls the distance-decay rate of the spatial correlation. One example
of a spatial covariance matrix is the ``exponential'', which is given by
\begin{equation}\label{eq-spcov-exp}
  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \exp(-\mathbf{H}/\phi),
\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
among the elements of \(\mathbf{y}\) and \(\phi\) is a range parameter.
From Equation\(~\)\ref{eq-spcov-exp}, as the distance between two
elements of \(\mathbf{y}\) increases, the spatial covariance decreases,
which reflects intuition. Moreover, as the range parameter, \(\phi\),
increases, the strength of spatial dependence increases
(Figure\(~\)\ref{fig-range}). The random effect
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
independent random errors. We assume that
\(\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}\), where
\(\mathbf{I}\) is an \(n \times n\) identity matrix. The variance
parameter \(\sigma^2_\epsilon\) controls the magnitude of nonspatial
variability (i.e., fine-scale variation) and is often called a nugget.

@@

Through inclusion of \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
\begin{equation}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{equation} Often in spatial statistics, quantities are explicitly
referenced with respect to \(\mathbf{s}\), a vector of coordinates
indexing the observation \citep{cressie1993statistics}. For example,
\(\mathbf{y}\) and \(\mathbf{X}\) may instead be written
\(\mathbf{y}(\mathbf{s})\) and \(\mathbf{X}\), respectively. We
acknowledge the utility of this nomenclature but drop the explicit
dependence on \(\mathbf{s}\) for simplicity of notation. Assuming
independence among \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\),
it follows that \begin{equation}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{equation} To better align with intuition, we henceforth
\(\sigma^2_{\tau}\) as \(\sigma^2_{de}\) (for spatially dependent error
variance) and \(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for
independent error variance). The parameters \(\sigma^2_{de}\),
\(\sigma^2_{ie}\), the range parameter \(\phi\) in \(\mathbf{R}\), and
any other parameters in \(\mathbf{R}\) compose \(\boldsymbol{\theta}\),
the covariance parameter vector.

Fitting and using SPGLMs is challenging both conceptually and
computationally \citep{bolker2009generalized}. Recently, however, there
have been numerous, significant advances in \proglang{R} software that
have made these models more accessible to practitioners. The \pkg{brms}
\citep{burkner2017brms}, \pkg{carBayes} \citep{lee2013carbayes},
\pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
\citep{lindgren2015bayesian} and \pkg{inlabru} \citep{bachl2019inlabru},
\pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
\citep{doser2022spoccupancy}, \pkg{spAbundance}
\citep{doser2024spabundance}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, can be computationally expensive. In order
to reduce computation time, many of these packages work with the
precision matrix instead of the covariance matrix so that
computationally expensive matrix inversion is not required. For example,
\pkg{R-INLA} uses the precision matrix and tends to be very fast.
Working with precision matrices, however, can be more restrictive and
less intuitive than working directly with the covariance matrix. The
{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
\citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
\pkg{mgcv} \citep{wood2017generalized}, and \pkg{spaMM}
\citep{rousset2014spamm} packages directly use Laplace,
quasi-likelihood, or reduced-rank approaches to estimate parameters.
These direct approaches tend to be computationally efficient, as they
don't rely on MCMC sampling. In contrast to the Bayesian approach, a
drawback of these direct approaches is that prior information cannot be
formally incorporated and covariance parameter uncertainty is more
challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and properties
of Gaussian Markov random fields to fit a wide variety of SPGLMs, and
\pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
to multivariate or (dynamic) structural equation models.

\citet{ver2024marginal} proposed a novel approach to fitting SPGLMs that
leverages the Laplace approximation while marginalizing over both the
latent \(\mathbf{w}\) and the fixed effects (\(\boldsymbol{\beta}\)) and
accommodating spatial covariance. \citet{ver2024marginal} showed that
this approach performed efficiently in a variety of simulation settings,
generally having appropriate confidence interval coverage for the fixed
effects and prediction interval coverage for new \(\mathbf{w}\). The
approach performed similarly to the Bayesian SPGLM approach in
\pkg{spBayes} and the automatic differentiation SPGLM approach in
\pkg{glmmTMB} but was much faster. At small sample sizes, the approach
outperformed the approximate Bayesian SPGLM approach in \pkg{R-INLA} and
had similar computational times. For moderate sample sizes, it performed
similarly to \pkg{R-INLA}, though \pkg{R-INLA} was faster. This novel
approach is particularly attractive for two reasons. First, it is
general enough that can be applied to any covariance structure (not just
spatial). Second, after estimating the covariance parameters, analytical
solutions exist for the fixed effects (and their standard errors) as
well as predictions of the latent \(\mathbf{w}\) at new locations (and
their standard errors). The \pkg{spmodel} \proglang{R} package
\citep{dumelle2023spmodel} recently provided full support for the
methods in \citet{ver2024marginal} applied to binary, count, skewed, and
proportion data for over 20 different spatial covariance types.

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently provided a full set of modeling tools for SPGLMs fit using the
methods described in \citet{ver2024marginal}. These modeling tools are
approachable and mirror the familiar \code{glm()} syntax from
base-\proglang{R}, making the transition from GLMs to SPGLMs relatively
seamless. The \code{spglm()} function fits SPGLMs for point-referenced
data (e.g., x-coordinates and y-coordinates representing point locations
in a field), while the \code{spgautor()} function fits SPGLMs for areal
data (e.g., polygon boundaries representing geographic subsets of a
region). \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
conditional autoregressive, and simultaneous autoregressive structures
for areal data. \pkg{spmodel} provides tools for commonly used model
summaries, visualizations, and diagnostics (e.g., Cook's distance) using
standard \proglang{R} helper functions like \code{summary()},
\code{plot()}, and \code{cooks.distance()}. \pkg{spmodel} also provides
tools to predict \(\mathbf{w}\) at new locations and quantify
uncertainty in those prediction using \code{predict()}. This core
functionality, combined with several advanced features we describe
throughout the manuscript, enables \pkg{spmodel} to provide some novel
and important capabilities previously missing from the existing SPGLM
ecosystem in \proglang{R}.

@@

\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB}
(version 0.7.4) in terms of scope and feel. Both packages use similar
syntax as \code{glm()}, accommodate flexible \code{formula} arguments
(e.g., offsets, splines), handle spatial covariance that decays at
different rates in different directions (i.e., geometric anisotropy),
incorporate nonspatial random effects, support other \proglang{R}
packages for modeling like \pkg{broom}
\citep{robinson2021broom, kuhn2022tidy}, \pkg{emmeans}
\citep{lenth2024emmeans}, and \pkg{car} \citep{fox2019car}, and have
tools for model summaries, prediction, and simulating data. There are
some notable differences between the two packages, however. \pkg{sdmTMB}
supports several additional GLM distributions like the Tweedie, supports
Hurdle models, and can incorporate prior information through Bayesian
applications. \pkg{sdmTMB} also provides tools for working with temporal
data and spatiotemporal data and provides enhanced visualizations of the
model's marginal effects. \pkg{sdmTMB} does require a preprocessing step
of constructing a mesh for the stochastic partial differential equation
approach, and the density of the mesh can affect model results and
computational complexity. On the other hand, \pkg{spmodel} does not
require the construction of a mesh prior to modeling. \pkg{spmodel}
supports 20 different spatial covariances and models them directly,
rather than using a precision matrix approximation to the Matérn spatial
covariance as in \pkg{sdmTMB}. \pkg{spmodel} can also model areal data
directly using neighborhood distance and autoregressive models rather
than relying on the polygon centroid (as in \pkg{sdmTMB}). \pkg{spmodel}
also provides experimental design tools (e.g., analysis of variance,
contrasts), supports \pkg{sf} objects in modeling and prediction
functions \citep{pebesma2018sf}, has several specialized model
diagnostics like leverage values and Cook's distances, and has analytic
solutions for prediction standard errors. Other similarities and
differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both
packages continue to evolve. Overall, we believe that these packages are
complementary and enhance the suite of SPGLM tools accessible to
practitioners.

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section\(~\)\ref{sec-applications}, we provide several applications of
\pkg{spmodel} to spatial binary, count, skewed and proportion data with
both point-referenced and areal supports. And in
Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.

\section{The spatial generalized linear model and
marginalizatoin}\label{sec-spglm}

\pkg{spmodel} implements the novel methods described in
\citet{ver2024marginal} to fit SPGLMs, which leverages the Laplace
approximation and marginalizes over both the latent \(\mathbf{w}\) and
the fixed effects while accommodating spatial covariance. A beneficial
aspect of this approach is that it formally maximizes a hierarchical GLM
likelihood \citep{lee1996hierarchical, wood2017generalized}. This makes
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not available for quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. \citet{ver2024marginal} provides
thorough details regarding the method and contextualizes its development
which built upon similar methods
\citep[\citet{bonat2016practical}]{evangelou2011estimation}. Next, we
describe a brief overview of the approach and how it can be used for
parameter estimation, inference, and prediction.

@@

\end{equation} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is
the density for the appropriate response distribution of \(\mathbf{y}\)

@@

\(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-spglm}, which yields
\begin{equation}\label{eq-marginal2}

@@

\end{equation} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\)
is the restricted (i.e., residual) multivariate Gaussian density

@@

be synonymous written after profiling the overall variance out of

@@

density is given by \begin{equation}\label{eq-reml-def}

@@

\end{equation} where

@@

\begin{equation}\label{eq-marginal03}

@@

\end{equation} and rewrite Equation\(~\)\ref{eq-marginal2} as
\begin{equation}\label{eq-marginal3}

@@

\end{equation} A second-order Taylor series expansion of

>>>>

\Abstract{
Non-Gaussian data are common in practice and include binary, count,
skewed, and proprtion data types. Often, non-Gaussian data are modeled
using a generalized linear model (GLM). GLMs typically assume that
observations are independent of one another. This is an impractical
assumption for spatial data, as nearby observations tend to be more
similar than distant ones. The \pkg{spmodel} package in \proglang{R}
provides a suite of tools for fitting spatial generalized linear models
(SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
Kriging). SPGLMs for point-referenced data (x- and y-coordinates) are
fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
polygon) data are fit using the \code{spgautor()} function. Both
\code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
which marginalizes over the model's fixed effects and latent mean while
formally incorporating spatial covariance (i.e., spatial correlation).
The inputs and outputs of \code{spglm()} and \code{spgautor()} closely
resemble the \code{glm()} function from base \proglang{R}, easing the
transition from GLMs to SPGLMs. \pkg{spmodel} provides and builds upon
several commonly used helper functions for model building like
\code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()},
among others. Spatial predictions of the latent mean at unobserved
locations are obtained using \code{predict()} or \code{augment()}.
\pkg{spmodel} accommodates myriad advanced modeling features like
geometric anisotrpoy, nonspatial random effects, analysis of variance,
and more. Throughout, we use \pkg{spmodel} to fit SPGLMs to moose
presence and counts in Alaska, United States (US), skewed conductivity
data in the Southwestern US, harbor seal abundance trends in Alaska, US,
and voter turnout rates in Texas, US.
}

@@

\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm, bbm}
\usepackage{lineno}
\usepackage{caption, subcaption}
\usepackage{autonum}

@@

\linenumbers

@@

In practice, non-Gaussian data (e.g., binary, count, skewed, and
proportion data) are ubiquitous. Non-Gaussian that belong to an
exponential family data can be naturally modeled using a generalized
linear model (GLM) regression framework
\citep{nelder1972generalized, mccullagh1989generalized}. In a GLM, an
\(n \times 1\) response variable \(\mathbf{y}\) belongs to a statistical
distribution (e.g., Binomial, Poisson) with some mean and variance.
Often, the analysis goal is to study the impact of a linear function of
several explanatory variables on \(\mathbf{y}\) through a GLM. In this
context, the latent (i.e., unobserved) mean of \(\mathbf{y}\),
\(\boldsymbol{\mu}\), is linked to these explanatory variables via a
link function: \begin{align}\label{eq-glm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{align} where for a sample size \(n\), \(f(\cdot)\) is a link
function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
\(\mathbf{X}\) is the \(n \times p\) design matrix of explanatory
variables, and \(\boldsymbol{\beta}\) is the \(p \times 1\) vector of
fixed effects. While the mean is typically constrained in some way
(e.g., if a probability, between zero and one), the link function
generally makes \(\mathbf{w}\) unconstrained. Common link functions
include the log odds (i.e., logit) link for binary and proportion data
and the log link for count and skewed data. Equation\(~\)\ref{eq-glm}
can also be written in terms of the inverse link function,
\(f^{-1}(\cdot)\): \begin{align}\label{eq-glm2}
\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}).
\end{align}

@@

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
For spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, which leads to positive
spatial covariance. The consequences of ignoring spatial covariance in
statistical models for spatial data can be severe and include imprecise
parameter estimates as well as misleading standard errors that inflate
Type-I error rates and decrease power \citep{zimmerman2024spatial}.

An approach for handling spatial data using a GLM is to assume the
elements of \(\mathbf{w}\) share covariance that varies spatially. This
is achieved by adding to Equation\(~\)\ref{eq-glm} two random effects,
\(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\). The random effect
\(\boldsymbol{\tau}\) is an \(n \times 1\) column vector of spatially
dependent random errors. We assume that
\(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\text{E}(\cdot)\) and \(\text{Cov}(\cdot)\) denote expectation and
covariance, respectively. The variance parameter \(\sigma^2_\tau\)
controls the magnitude of spatial covariance and is often called a
partial sill. The matrix \(\mathbf{R}\) is an \(n \times n\) spatial
correlation matrix that depends on a range parameter controlling the
distance-decay rate of the spatial correlation. One example of a spatial
covariance matrix is the ``exponential'', which is given by
\begin{align}\label{eq-spcov-exp}
  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \mathbf{R}_{exp} =  \sigma^2_{de} \exp(-\mathbf{H}/\phi),
\end{align} where \(\mathbf{H}\) is a matrix of pairwise distances among
the elements of \(\mathbf{y}\) and \(\phi\) is the range parameter. From
Equation\(~\)\ref{eq-spcov-exp}, as the distance between two elements of
\(\mathbf{y}\) increases, the spatial covariance decreases, which
reflects intuition. Moreover, as the range parameter, \(\phi\),
increases, the strength of spatial dependence increases
(Figure\(~\)\ref{fig-range}). The random effect
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
independent random errors. We assume that
\(\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}\), where
\(\mathbf{I}\) is an \(n \times n\) identity matrix. The variance
parameter \(\sigma^2_\epsilon\) controls the magnitude of nonspatial
variability (i.e., fine-scale variation) and is often called a nugget.
Often in spatial statistics, quantities are explicitly referenced with
respect to \(\mathbf{s}\), a vector of coordinates indexing the
observation \citep{cressie1993statistics}. For example, \(\mathbf{y}\)
and \(\mathbf{X}\) may instead be written \(\mathbf{y}(\mathbf{s})\) and
\(\mathbf{X}\), respectively. We acknowledge the utility of this
nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
simplicity of notation.

@@

Through inclusion of \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
\begin{align}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{align} Assuming independence among \(\boldsymbol{\tau}\) and
\(\boldsymbol{\epsilon}\), it follows that \begin{align}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{align} Henceforth, we refer to \(\sigma^2_{\tau}\) as
\(\sigma^2_{de}\) (for spatially dependent error variance) and
\(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), and
\(\phi\), in addition to any other parameters in \(\mathbf{R}\), compose
\(\boldsymbol{\theta}\), the covariance parameter vector.

Fitting and using SPGLMs is challenging both conceptually and
computationally \citep{bolker2009generalized}. Recently, however, there
have been numerous, significant advances in \proglang{R} software that
have made these models more accessible to practitioners. The \pkg{brms}
\citep{burkner2017brms}, \pkg{carBayes} \citep{lee2013carbayes},
\pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
\citep{lindgren2015bayesian} and \pkg{inlabru} \citep{bachl2019inlabru},
\pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
\citep{doser2022spoccupancy}, \pkg{spAbundance}
\citep{doser2024spabundance}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, can be computationally expensive. In order
to reduce computation time, many of these packages work with the
precision matrix instead of the covariance matrix so that
computationally expensive matrix inversion is not required. For example,
\pkg{R-INLA} uses the precision matrix and tends to be very fast.
Working with precision matrices, however, can be more restrictive and
less intuitive than working directly with the covariance matrix. The
\pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
\citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
\pkg{mgcv} \citep{wood2017generalized}, and \pkg{spaMM}
\citep{rousset2014spamm} packages directly use Laplace,
quasi-likelihood, or reduced-rank approaches to estimate parameters.
These direct approaches tend to be computationally efficient, as they
don't rely on MCMC sampling. In contrast to the Bayesian approach, a
drawback of these direct approaches is that prior information cannot be
formally incorporated and covariance parameter uncertainty is more
challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and Gaussian
Markov random fields to fit a wide variety of SPGLMs, while
\pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
to multivariate or (dynamic) structural equation models.

Building from \citet{evangelou2011estimation} and
\citet{bonat2016practical}, \citet{ver2024marginal} proposed a novel
approach for fitting SPGLMs that leverages the Laplace approximation
while marginalizing over both the latent \(\mathbf{w}\) and the fixed
effects (\(\boldsymbol{\beta}\)) and accommodating spatial covariance.
This approach performed efficiently in a variety of simulation settings,
generally having appropriate confidence interval coverage for the fixed
effects and prediction interval coverage for \(\mathbf{w}\) at new
locations. The approach performed similarly to the Bayesian SPGLM
approach in \pkg{spBayes} and the automatic differentiation SPGLM
approach in \pkg{glmmTMB} but was much faster. At small sample sizes,
the approach outperformed the approximate Bayesian SPGLM approach in
\pkg{R-INLA} and had similar computational times. For moderate sample
sizes, it performed similarly to \pkg{R-INLA}, though \pkg{R-INLA} was
faster. The novel Laplace approach is particularly attractive for two
reasons. First, it is general enough that can be applied to any
covariance structure (not just spatial). Second, after estimating the
covariance parameters, analytical solutions exist for the fixed effects
(and their standard errors) as well as predictions of the latent
\(\mathbf{w}\) at new locations (and their standard errors).

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently provided a full set of modeling tools for SPGLMs fit using the
novel Laplace approach described by \citet{ver2024marginal}. These
modeling tools are approachable and mirror the familiar \code{glm()}
syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs
relatively seamless. The \code{spglm()} function fits SPGLMs for
point-referenced data (e.g., x-coordinates and y-coordinates
representing point locations in a field; models are sometimes called
``geostatistical'' models), while the \code{spgautor()} function fits
SPGLMs for areal data (e.g., polygon boundaries representing geographic
subsets of a region; models are sometimes called ``autoregressive''
models). \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
conditional autoregressive, and simultaneous autoregressive structures
for areal data. \pkg{spmodel} provides tools for commonly used model
summaries, visualizations, and diagnostics (e.g., Cook's distance) using
standard \proglang{R} helper functions like \code{summary()},
\code{plot()}, \code{fitted()}, and \code{tidy()}, among othres.
\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
locations and quantify uncertainty in those prediction using
\code{predict()} and \code{augment()}. This core functionality, combined
with several advanced features we describe throughout the manuscript,
enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
previously missing from the existing \proglang{R} ecosystem.

@@

\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB}
(version 0.7.4) in terms of scope and feel. Both packages use similar
syntax as \code{glm()}, accommodate flexible \code{formula} arguments
(e.g., offsets, splines), handle spatial covariance that decays at
different rates in different directions (i.e., geometric anisotropy),
incorporate nonspatial random effects, support other \proglang{R}
packages for modeling like \pkg{broom}
\citep{robinson2021broom, kuhn2022tidy}, \pkg{emmeans}
\citep{lenth2024emmeans}, and \pkg{car} \citep{fox2019car}, and have
tools for model summaries, prediction, and simulating data. There are
some notable differences between the two packages, however. \pkg{sdmTMB}
supports several additional GLM distributions like the Tweedie, supports
Hurdle models, and can incorporate prior information through Bayesian
applications. \pkg{sdmTMB} also provides tools for working with temporal
data and spatiotemporal data and provides enhanced visualizations of the
model's marginal effects. \pkg{sdmTMB} does require a preprocessing step
of constructing a mesh for the stochastic partial differential equation
approach, and the density of the mesh can affect model results and
computational complexity. On the other hand, \pkg{spmodel} does not
require the construction of a mesh prior to modeling. \pkg{spmodel} also
supports 20 different spatial covariances and models them directly,
rather than using a precision matrix approximation to the Matérn spatial
covariance as in \pkg{sdmTMB}. \pkg{spmodel} can also model areal data
directly using neighborhood distance and autoregressive models rather
than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not
be within the polygon's boundaries. \pkg{spmodel} also provides
experimental design tools (e.g., analysis of variance, contrasts),
supports \pkg{sf} objects in modeling and prediction functions
\citep{pebesma2018sf}, has several specialized model diagnostics like
leverage values and Cook's distances, and has analytic solutions for
fixed effect and prediction standard errors. Other similarities and
differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both
packages continue to evolve. Overall, we believe that these packages are
complementary and enhance the suite of SPGLM tools accessible to
practitioners.

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section\(~\)\ref{sec-applications}, we provide an overview of core SPGLM
functionaility in \pkg{spmodel} by modeling moose presence in Alaska,
United States (US). In Section\(~\)\ref{sec-applications2}, we model
moose counts in Alaska, US; skewed lake conductivity in the Southwestern
US; harbor seal abundance trend behavior in Alaska, US; and voter
turnout rates in Texas, US. And in Section\(~\)\ref{sec-discussion}, we
end with a discussion synthesizing \pkg{spmodel}'s contributions to the
analysis of SPGLMs in \proglang{R}.

\section{The spatial generalized linear model and
marginalization}\label{sec-spglm}

The novel Laplace approach implemented in \pkg{spmodel} formally
maximizes a hierarchical GLM likelihood
\citep{lee1996hierarchical, wood2017generalized}, making
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not available for quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. Next, we describe a brief overview
of the approach and how it can be used for several primary data analysis
tasks \citep{tredennick2021practical} like model comparison, parameter
estimation, inference, model diagnostics, and prediction.

@@

\end{align} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is the
density for the appropriate response distribution of \(\mathbf{y}\)

@@

\(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-marginal}, which
yields \begin{align}\label{eq-marginal2}

@@

\end{align} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is
the restricted (i.e., residual) multivariate Gaussian density

@@

synonymously be written after profiling the overall variance out of

@@

density is given by \begin{align}\label{eq-reml-def}

@@

\end{align} where

@@

\begin{align}\label{eq-marginal03}

@@

\end{align} and rewrite Equation\(~\)\ref{eq-marginal2} as
\begin{align}\label{eq-marginal3}

@@

\end{align} A second-order Taylor series expansion of
\(\ell_\mathbf{w}\) around a point \(\hat{\mathbf{w}}\) yields
\begin{align}\label{eq-marginal4}
